---
title: "Pracitcal Machine Learning - Course Project"
author: "Senthil Kumar K"
date: "Tuesday, December 16, 2014"
output: html_document
---

===========================================

This is a submission for Coursera: Practical Machine Learning- Course Project

##Background##

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

##Data##

The training data for this project are available [here](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv)

The test data are available [here](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv)

The data for this project come from this [source](http://groupware.les.inf.puc-rio.br/har). If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

##Reproduceability##

An overall pseudo-random number generator seed was set at 12345 for all code. In order to reproduce the results below, the same seed should be used.
Different packages were downloaded and installed, such as caret and randomForest. These should also be installed in order to reproduce the results below (please see code below for ways and syntax to do so).

###How the model was built###

Our outcome variable is classe, a factor variable with 5 levels. For this data set, "participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in 5 different fashions:

- exactly according to the specification (Class A)
- throwing the elbows to the front (Class B)
- lifting the dumbbell only halfway (Class C)
- lowering the dumbbell only halfway (Class D)
- throwing the hips to the front (Class E)

Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes." [1]
Prediction evaluations will be based on maximizing the accuracy and minimizing the out-of-sample error. All other available variables after cleaning will be used for prediction.
Two models will be tested using decision tree and random forest algorithms. The model with the highest accuracy will be chosen as our final model.

###Cross-validation###

Cross-validation will be performed by subsampling our training data set randomly without replacement into 2 subsamples: subTraining data (75% of the original Training data set) and subTesting data (25%). Our models will be fitted on the subTraining data set, and tested on the subTesting data. Once the most accurate model is choosen, it will be tested on the original Testing data set.

###Expected out-of-sample error###

The expected out-of-sample error will correspond to the quantity: 1-accuracy in the cross-validation data. Accuracy is the proportion of correct classified observation over the total sample in the subTesting data set. Expected accuracy is the expected accuracy in the out-of-sample data set (i.e. original testing data set). Thus, the expected value of the out-of-sample error will correspond to the expected number of missclassified observations/total observations in the Test data set, which is the quantity: 1-accuracy found from the cross-validation data set.

##Coding and Results##

**Packages, Libraries and Seed**

Load all the required packages and libraries into R.
Set seed for reproducibility

```{r}
library(caret)
library(rattle)
library(rpart)
library(randomForest)
set.seed(12345)
```

**Load Data**

We first load the dat into R and make sure the missing values are coded properly.
Some missing values are coded as string "#DIV/0!" or "" or "NA" - these will be changed to NA.

```{r}
trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

training <- read.csv(url(trainUrl), na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(url(testUrl), na.strings=c("NA","#DIV/0!",""))
```

**Partition the data into Training and Test sets**
Partioning Training data set into two data sets, 70% for myTraining, 30% for myTesting

```{r}
inTrain <- createDataPartition(y=training$classe, p=0.7, list=FALSE)
myTraining <- training[inTrain, ]
myTesting <- training[-inTrain, ]
dim(myTraining)
dim(myTesting)
```

**Data Cleansing**

Data is transformed in following 3 steps:  
1.Cleaning NearZeroVariance Variables.  
2.Removing first column of Dataset - so that it does not interfer with ML Algorithms.  
3.Cleaning Variables with too many NAs. Leave out Variables that have more than a 60% threshold of NA's.  

```{r}
myDataNZV <- nearZeroVar(myTraining, saveMetrics=TRUE)
myNZVvars <- names(myTraining) %in% c("new_window", "kurtosis_roll_belt", "kurtosis_picth_belt",
                                      "kurtosis_yaw_belt", "skewness_roll_belt", "skewness_roll_belt.1", "skewness_yaw_belt",
                                      "max_yaw_belt", "min_yaw_belt", "amplitude_yaw_belt", "avg_roll_arm", "stddev_roll_arm",
                                      "var_roll_arm", "avg_pitch_arm", "stddev_pitch_arm", "var_pitch_arm", "avg_yaw_arm",
                                      "stddev_yaw_arm", "var_yaw_arm", "kurtosis_roll_arm", "kurtosis_picth_arm",
                                      "kurtosis_yaw_arm", "skewness_roll_arm", "skewness_pitch_arm", "skewness_yaw_arm",
                                      "max_roll_arm", "min_roll_arm", "min_pitch_arm", "amplitude_roll_arm", "amplitude_pitch_arm",
                                      "kurtosis_roll_dumbbell", "kurtosis_picth_dumbbell", "kurtosis_yaw_dumbbell", "skewness_roll_dumbbell",
                                      "skewness_pitch_dumbbell", "skewness_yaw_dumbbell", "max_yaw_dumbbell", "min_yaw_dumbbell",
                                      "amplitude_yaw_dumbbell", "kurtosis_roll_forearm", "kurtosis_picth_forearm", "kurtosis_yaw_forearm",
                                      "skewness_roll_forearm", "skewness_pitch_forearm", "skewness_yaw_forearm", "max_roll_forearm",
                                      "max_yaw_forearm", "min_roll_forearm", "min_yaw_forearm", "amplitude_roll_forearm",
                                      "amplitude_yaw_forearm", "avg_roll_forearm", "stddev_roll_forearm", "var_roll_forearm",
                                      "avg_pitch_forearm", "stddev_pitch_forearm", "var_pitch_forearm", "avg_yaw_forearm",
                                      "stddev_yaw_forearm", "var_yaw_forearm")

myTraining <- myTraining[!myNZVvars]
dim(myTraining)

myTraining <- myTraining[c(-1)]

trainingV3 <- myTraining 
for(i in 1:length(myTraining)) { 
  if( sum( is.na( myTraining[, i] ) ) /nrow(myTraining) >= .6 ) { 
    for(j in 1:length(trainingV3)) {
      if( length( grep(names(myTraining[i]), names(trainingV3)[j]) ) ==1)  { 
        trainingV3 <- trainingV3[ , -j]
      }  
    } 
  }
}
dim(trainingV3)

myTraining <- trainingV3
rm(trainingV3)
```

**A look at the data**

The variable "classe" contains 5 levels: A, B, C, D and E. A plot of the outcome variable will allow us to see the frequency of each levels in the subTraining data set and compare one another.

```{r}
plot(myTraining$classe, col="blue", main="Bar Plot of levels of the variable classe within the myTraining data set", xlab="classe levels", ylab="Frequency")
```

Now, we do the same transformation on the training sets

```{r}
clean1 <- colnames(myTraining)
clean2 <- colnames(myTraining[, -58])
myTesting <- myTesting[clean1]
testing <- testing[clean2]

dim(myTesting)
dim(testing)
```

**Coerce data**

In order to ensure proper functioning of Decision Trees, we need to coerce the data into the same type.

```{r}
for (i in 1:length(testing) ) 
{
  for(j in 1:length(myTraining)) 
  {
    if( length( grep(names(myTraining[i]), names(testing)[j]) ) ==1)  
    {
      class(testing[j]) <- class(myTraining[i])
    }      
  }      
}

testing <- rbind(myTraining[2, -58] , testing) 
testing <- testing[-1,]
```

###First prediction model: Using Decision Tree###

**Use ML algorithms for prediction**
```{r}
modFitA1 <- rpart(classe ~ ., data=myTraining, method="class")
```

**Predicting**
```{r}
predictionsA1 <- predict(modFitA1, myTesting, type = "class")
```

**Plot a Decision Tree**
```{r}
fancyRpartPlot(modFitA1)
```

**Use Confusion Matrix to Test the Results**
```{r}
confusionMatrix(predictionsA1, myTesting$classe)
```

###Second prediction model: Using Random Forest###

**We again use ML algorithms for prediction**
```{r}
modFitB1 <- randomForest(classe ~. , data=myTraining)
```

**Predict**
```{r}
predictionsB1 <- predict(modFitB1, myTesting, type = "class")
```

**Use Confusion matrix to test the results**
```{r}
confusionMatrix(predictionsB1, myTesting$classe)
```

##Conclusion##

**As expected, Random Forest algorithm performed better than Decision Trees.**
Accuracy for Random Forest model was 0.9988 (95% CI: (0.9976, 0.9995)) compared to 0.8724 (95% CI: (0.8636, 0.8808)) for Decision Tree model. 
The random Forest model is choosen. The accuracy of the model is 0.9988. The expected out-of-sample error is estimated at 0.005, or 0.5%. The expected out-of-sample error is calculated as 1 - accuracy for predictions made against the cross-validation set. Our Test data set comprises 20 cases. With an accuracy above 99% on our cross-validation data, we can expect that very few, or none, of the test samples will be missclassified.

##Submission##

Predict outcome levels on the original Testing data set using Random Forest algorithm

```{r}
predictionsB2 <- predict(modFitB1, testing, type = "class")
predictionsB2
```

**Generate Files for the Assignment**
Function to generate files for assignment
```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predictionsB2)
```
